Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-09-02 papers/paper/ms.tex
update ms

\received{2024 June 21}
\revised{2024 September 5}
\accepted{---}
\shorttitle{Kepler's Demographic Cliff}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}
\affiliation{\carnegie}

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{{\bf Potential impact of unresolved binaries.}   This figure shows
		the same raw (thinnest line) and quality-flag cleaned (``gyro applicable'') distributions
		as in Figure~\ref{fig:hist_tgyro}.
		However, even after applying our quality cuts, up to
		\fivepctnuniqstarsantosrotgyroappl\ of \ngyroappllttendays\ ``clean'' rotators with
		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
		The thick violet ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
		what we estimate (estm.) to be the maximum possible number of such systems; the true
		correction might be smaller.  We expect such a correction to be more important in
		the sample of all Kepler stars than in the KOI host sample.
		\label{fig:hist_tgyro_corr}
	}
\end{figure*}

@@

\added{{\it Rotational outliers:} }
We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for \replaced{young
stars known to host planets}{each star in our sample}.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, similar to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  \added{While most of these outliers are known to have bad quality flags, some do not.} 
These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found consistent $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
($\approx$4.4\,Gyr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

@@

\added{Other age indicators may help in verifying the ages of the
$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
can serve as an age tracer
\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
However, these age indicators are all in a sense ``rotation-powered''.
The dynamo converts kinetic energy into magnetic energy, which is
emitted through these chromospheric pathways.  We did not attempt to
incorporate these indicators into this study due to concerns regarding
sensitivity, homogeneity, and the question of whether they in fact
provide age information that is truly independent from rotation.}

@@

\added{{\it Metallicity:}  
Stars with higher metallicity are thought to have longer convective turnover
times, which may slow their rate of spin-down
	\citep{Amard2020}.
While observational studies of this effect have been suggestive, further work
remains to definitively establish the importance of the effect \citep[e.g.][]{Amard2020b,2024arXiv240500779S}
Nonetheless, based on the theoretical predictions, we flagged stars with known
	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, based on either LAMOST or Keck/HIRES
	spectra \citep{Zong2018,Petigura_2022}.
The main limitation of this approach is that only 53.3\% of the stars in Table~\ref{tab:stars} with
$t_{\rm gyro}$$<$4\,Gyr had such spectra available.
For the rest of the sample, primarily composed of fainter stars,
this means that our quality flags were not sensitive to metallicity outliers.
Based on the metallicity distribution of stars for which we do have spectra,
this 
implies that at most $\approx$5\% of stars in our sample might have $|[{\rm Fe/H}]|$$>$0.3,
with a slight bias toward lower metallicity stars over higher metallicity.
Based on the results by \citet{Claytor2020},  this will be most important
for stars near the Kraft break, which could have ages over-estimated
by $\approx$50-100\% if they are in fact metal-poor, and under-estimated in the converse case.
}

\added{{\it Differential rotation:}
The Sun's surface rotates once per $\approx$25-30\,days at latitudes between the equator
and $\approx$$\pm$60$^\circ$ \citep{Snodgrass1990}.
This latitudinal surface differential rotation yields two separate limitations on rotational ages.
The first is
on the accuracy with which any one star's rotation period can be measured over a few-year mission like Kepler;
the second is on the degree to which the intrinsic scatter in open cluster rotation sequences can be measured \citep[e.g.][]{Epstein_2014}.
In this study, we accounted for the latter intrinsic scatter based on available measurements
from Ruprecht-147 \citep{Curtis_2020} and M67 data
\citep{Barnes2016,Dungee_2022,Gruner_2023}.
This means that irrespective of the assumed $P_{\rm rot}$ measurement precision,
our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter (perhaps differential-rotation induced)
at old ages.
While this accounts for the statistical impact of differential rotation on the inferred
age distribution,
the accuracy issue remains;
any one individual star's observed rotation period might be faster
or slower than average stars of the same mass and age.  
This effect can yield a $\approx$20\% age bias for individual Sun-like stars with ages
of a few billion years; the statistical uncertainties in Table~\ref{tab:stars} should be
treated with appropriate caution.
}

@@

\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
old.}\added{This work was limited to stars younger than 4\,Gyr}. For
age-dating methods based only on rotation rates, an important challenge
is that the detectability of the rotation signals for 3-10\,Gyr stars,
especially Sun-like stars, is at the limits of the Kepler data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
the evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
capable of providing useful age constraints for single stars over the
full span of their main-sequence lifetimes.

{\it \added{Planet} occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.  Some care is required since
beyond age, exoplanet demographics also depend on stellar metallicity
and mass \citep[e.g.][]{Petigura_2018,Miyazaki2023}.

{\it Field star ages from other surveys}:
While this study focused on Kepler, other photometric surveys
(e.g.~K2, TESS, HAT, WASP, NGTS, ZTF, ATLAS) open opportunities for
age-dating a far broader set of stars and planets.  Future prospects
also include PLATO \citep{Rauer14}, Earth 2.0
\citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
Survey \citep{Wilson2023}.  Rotation-based age-dating from these surveys could
yield new insight
into whether the star formation history that we see in the Kepler field is
universal across the thin disk.

@@

\added{Significant care is 
needed not only when assessing the reliability of individual rotational ages, but also when interpreting distributions of rotational ages inferred from many stars.
Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
and binarity, we expect that up to a third of our ``apparently clean'' 
Kepler stars with $P_{\rm rot}$$<$10\,days
could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
An additional nuance is that even after correcting for effects
that can bias rotational ages, our observed age distribution does not 
directly yield a star formation history: 
particularly for pencil-beam surveys such as Kepler,
kinematic heating of stars in the
thin disk significantly sculpts the stellar (and therefore planetary) age distribution.}

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.}
\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty in the contamination rate of unresolved
    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
     \replaced{This result}{This range of values} \replaced{from
    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} CMD fitting and white-dwarf
    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we \replaced{should}{might} instead expect $\approx$250.
  %
  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

>>>>

\received{2024 June 21}
\revised{2024 September 4}
\accepted{---}
\shorttitle{Kepler's Demographic Cliff}

@@

\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow;  Carnegie Fellow}
\affiliation{\caltech}
\affiliation{\carnegie}

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
		}
	\end{center}
	\vspace{-0.6cm}
	\caption{{\bf Potential impact of unresolved binaries.}   This figure shows
		the same raw (thinnest line) and quality-flag cleaned (``gyro applicable'') distributions
		as in Figure~\ref{fig:hist_tgyro}.
		However, even after applying our quality cuts, up to
		\fivepctnuniqstarsantosrotgyroappl\ of \ngyroappllttendays\ apparently clean rotators with
		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
		The black ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
		what we estimate (estm.) to be the maximum possible number of such systems; the true
		correction might be smaller.  We expect such a correction to be more important in
		the sample of all Kepler stars than in the KOI host sample.
		\label{fig:hist_tgyro_corr}
	}
\end{figure*}

@@

\added{{\it Rotational outliers:} }
We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for \replaced{young
stars known to host planets}{each star}.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were mostly accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, similar to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  \added{While most of these outliers have known bad quality flags, some do not.} 
These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found consistent $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
($\approx$4.4\,Gyr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

@@

%\added{Other age indicators may help in verifying the ages of the
%$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
%chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
%can serve as an age tracer
%\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
%However, these age indicators are all in a sense ``rotation-powered''.
%The dynamo converts kinetic energy into magnetic energy, which is
%emitted through these chromospheric pathways.  We did not attempt to
%incorporate these indicators into this study due to concerns regarding
%sensitivity, homogeneity, and the question of whether they in fact
%provide age information that is truly independent from rotation.}

@@

\added{{\it Metallicity:}  
Stars with higher metallicity are predicted to have longer convective turnover
times, which may slow their rate of spin-down
	\citep{Amard2020}.
While recent observational work has empirically confirmed that such an effect is
present in the Kepler data
\citep[][]{2024arXiv240500779S},
the anti-correlation between close,
unresolved binaries and stellar metallicity \citep{Moe2019} could be a complicating factor.
Nonetheless, based on these studies, we flagged stars with known
	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, using either LAMOST or Keck/HIRES
	spectra \citep{Zong2018,Petigura_2022}.
The main limitation of this approach is that only 53.3\% of the stars in Table~\ref{tab:stars} with
$t_{\rm gyro}$$<$4\,Gyr had such spectra available.
For the remainder of our sample,
our quality flags were not sensitive to metallicity outliers.
The metallicity distribution of stars for which we do have spectra suggests that
at most $\approx$5\% of stars in our apparently clean sample might have $|[{\rm Fe/H}]|$$>$0.3 and not
be flagged,
with a slight bias toward lower metallicity stars over higher metallicity.
Based on the study by \citet{Claytor2020},  this will be most important
for stars near the Kraft break, which could have ages over-estimated
by $\approx$50-100\% if they are in fact metal-poor, and under-estimated in the converse case.
}

\added{{\it Differential rotation:}
The Sun's surface rotates once per $\approx$25-30\,days at latitudes between the equator
and $\approx$$\pm$60$^\circ$ \citep[e.g.][]{Snodgrass1990}.
This latitudinal surface differential rotation yields two separate limitations on rotational ages.
The first is
on the accuracy with which any one star's rotation period can be measured over a few-year mission like Kepler;
the second is on the degree to which the intrinsic scatter in open cluster rotation sequences can be measured \citep[e.g.][]{Epstein_2014}.
In this study, we accounted for the latter intrinsic scatter based on available measurements
from Ruprecht-147 \citep{Curtis_2020} and M67 data
\citep{Barnes2016,Dungee_2022,Gruner_2023}.
This means that irrespective of the assumed $P_{\rm rot}$ measurement precision,
our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter
at old ages.
While this accounts for the statistical impact of differential rotation on the inferred
age distribution,
the accuracy issue remains;
any one individual star's observed rotation period might be faster
or slower than average stars of the same mass and age.  
This effect can yield a $\approx$20\% age error for individual Sun-like stars with ages
of a few billion years; the statistical uncertainties in Table~\ref{tab:stars} should be
treated with appropriate caution.}

@@

\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
old.}\added{This work was limited to stars younger than 4\,Gyr}. For
age-dating methods based only on rotation rates, an important challenge
is that the detectability of the rotation signals for 3-10\,Gyr stars,
especially Sun-like stars, is at the limits of the Kepler data
\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
leverage either asteroseismology
\citep{vanSaders_2016,2024ApJ...962..138S}, kinematics \citep{2021AJ....161..189L,2024arXiv240500779S},
or else that combine both
the evolution of stellar luminosity and stellar rotation
\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
capable of providing useful age constraints for single stars over the
full span of their main-sequence lifetimes.

{\it \added{Planet} occurrence rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.  Some care is required since
beyond age, exoplanet demographics also depend on stellar metallicity
and mass \citep[e.g.][]{Petigura_2018,Miyazaki2023}.

{\it Field star ages from other surveys}:
While this study focused on Kepler, other photometric surveys
(e.g.~K2, TESS, HAT, WASP, NGTS, ZTF, ATLAS) open opportunities for
age-dating a far broader set of stars and planets.  Future prospects
also include PLATO \citep{Rauer14}, Earth 2.0
\citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
Survey \citep{Wilson2023}.  Rotation-based ages from these surveys could
yield new insight
into \replaced{whether the star formation history that we see in the Kepler field is
universal across the thin disk.}{the relative importance of star formation and
kinematic heating across different components of the Galaxy.}

@@

\added{Significant care is 
needed not only when assessing the reliability of individual rotational ages, but also when interpreting distributions of rotational ages inferred from many stars.
Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
and binarity, we expect that up to a third of our ``apparently clean'' 
Kepler stars with $P_{\rm rot}$$<$10\,days
could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
An additional nuance is that even after correcting for such effects,
the star formation history does not directly trace
the observed age distribution: 
particularly for pencil-beam surveys such as Kepler,
kinematic heating of stars in the
thin disk significantly sculpts the observed stellar (and therefore planetary) age distribution.}

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.}
\item \added{The youngest Kepler stars (0-0.3\,Gyr)
    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty in the contamination rate of unresolved
    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
     \replaced{This result}{This range of values} \replaced{from
    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} CMD fitting and white-dwarf
    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
    individual FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 cumulative exoplanets younger than one
    billion years, the age distribution of the Kepler field
    \replaced{implies}{suggests} that we \replaced{should}{might} instead expect $\approx$250.
  %
  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}
